Classless Java

### 2016-03-21 OOPSLA16/sections/Overview.tex
Section 2

This section illustrates how our programming style, supported by the \mixinAnn{} annotation, enable powerful
idioms similar to multiple inheritance and field type refinement.
We propose a standard example: \Q@Animal@s with a two-dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horses@ and \Q@Bird@s.
Birds can \Q@fly@, thus their locations need to be three-dimensional \Q@Point3D@s (field type refinement).
Finally, we model \Q@Pegasus@ (one of the best-known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s (multiple inheritance). A simple class diagram illustrating
the basic system is given on the left side of Figure~\ref{fig:pegasus}.\footnote{Some research argues in favor of using subtyping for modeling taxonomies, other research argues against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}
\bruno{Should we provide a one sentence summary in the abstract of how much code
is needed in Java (without Obj) vs the approach with CJ, for the Pegasus example?}
\marco{Not in the abstract. But I think is good to have that number and to put in in the example section.
I will set up the sentence and someone can compute the number.}
\subsection{Simple Multiple Inheritance with Default
  Methods}\label{sec:simple}

@@

\lstinputlisting[linerange=8-10]{../UseMixinLombok/src/pegasus/simple/lombok/Main.java}% APPLY:linerange=PEGASUS_LOMBOK

\noindent The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a \texttt{Pegasus} object is instantiated as follows:

@@

The animal system modeled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}.
Here you still can not appreciate the advantage of IB.
Lets model now the complete animal system where \Q@Animal@s will include a \Q@location@
representing their positions in space. We will use 2D points to keep track of locations using coordinates.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Here we will illustrate how points are modelled with
interfaces. In IB we do not talk about state directly; state is accessed
using abstract (getter) methods. \bruno{unfinished sentence: Note how in}   The usual
approach to model points in Java is to use a class with fields for the
coordinates.

@@

\noindent Our encoding over Java is now inconvenient: creating a new point object is cumbersome, even
with anonymous classes:

\begin{lstlisting}
Point2D p = new Point2D() {
  public int x(){return 4;} public int y(){return 2;}  }
\end{lstlisting}

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with object interfaces}
This obvious ``constructor'' code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor. \mixin first looks at the abstract methods and detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{with-} methods in object interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable
data-structures. As such, it is
supported in our code generation as \Q@with-@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D {
    int x();    int y(); // getters
    Point2D withX(int val);    Point2D withY(int val); //with methods
}
\end{lstlisting}

@@

\noindent If there is a large number of fields, \texttt{with} methods
will save programmers from writing large amounts of tedious code that
simply copies field values.
\begin{comment}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provides a more uniform translation that can work also for mutable data structures, where setters are required.
}

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state:
\marcoT{Method bodies can only refer to other methods;
Objects are (conceptually) closures and the state is 
composed by captured local variables.}
Methods interacting with those are called \emph{abstract state operations}.

Object interfaces provide support for four
different types of abstract state operations \textcolor{red}{together with two types of
general operations}, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in more detail in Section~\ref{sec:dsls}.

\begin{figure*}
\centering
\saveSpaceFig
\begin{tabular}{|l|l|l|l|}
\hline
& \textbf{Operation}  & \textbf{Example}                  & \textbf{Description } \\ \hline
\multirow{4}{*}{State operations} & \textbf{``fields''/getters}        &   \Q@int x()@                  & Retrieves value from field \texttt{x}.          \\ \cline{2-4}
& {\bf withers}        &   \Q@Point2D withX(int val)@                & Clones
object; updates field \texttt{x} to \texttt{val}.             \\ \cline{2-4}
& \textbf{setters}        & \Q@void x(int val)@ & Sets the field
\texttt{x} to a  new value \texttt{val}.        \\ \cline{2-4}
& \textbf{fluent setters}        & \Q@Point2D x(int val)@ &Sets the field
\texttt{x} to \texttt{val} and returns \texttt{\this}.           \\ \hline
\multirow{3}{*}{Other operations} &
\textbf{factory methods} & 
%\begin{tabular}{l}
\Q@static Point2D of(int _x,int _y)@
%\end{tabular}
 & Generate factory method for \texttt{Point2D}.        \\
\cline{2-4}
& \textbf{functional updaters}        & \Q@Point3D with(Point2D val)@ & Update all fields
that match fields in \texttt{val}.        \\
\cline{2-4}
& \textbf{imperative updaters}        & \Q@Point3D set(Point2D val)@ & Set all fields
that match fields in \texttt{val}.        \\ \hline
\end{tabular}

\caption{\textcolor{red}{(edited2)} Abstract state operations for a field \texttt{x}, together with general operations, supported by the \mixin
  annotation. }

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the \Q@Animal@ interface, the notion of points needs to
be refined. Such kind of refinement is challenging
in typical class-based approaches. Fortunately, with object interfaces,
we are able to provide a simple and effective solution.

@@

%Interface \texttt{Point3D} extends \texttt{Point2D} with a new abstract method
%\texttt{int z()} (treated as a getter for member field \texttt{z}). Note that
%the return type of various methods (e.g. with- methods, getters) get refined
%either by covariant method overriding or automatically by our annotation
%processor. Besides \emph{with}, other methods (including \emph{clone},
%\emph{of}) also do type-refinements automatically.

@@

\bruno{At some point we should talk about the point that was discussed in the reviews}
\marco{you mean the reviewer question on the list.size?}
\paragraph{\Q@Point3D@ and properties updaters}
The \Q@Point3D@ interface is defined as follows:

@@

@@

 \marcoT{Symmetrically,
 we could offer an imperative property updater
 that calls the setters instead of the withers.
 \Q@Point3D set(Point2D val)@.
 }

 \marcoT{
 Property updaters never break class invariants, since they
 internally call operations that was already deemed
 safe by the programmer. For example a list object
 would not offer a setter for its size field, thus
 a property updater would not attempt to set it.}

@@

@@

\subsection{Our Approach}
Instead of writing a whole another class to provide the implementation for
\texttt{Point}, we annotate on interface \texttt{Point} directly with \mixin:

@@

With our approach, we provide a Java annotation \mixin to provide default
implementations for various methods and a mechanism to instantiate
objects. \mixin annotation helps programmers to write less cumbersome code and
instantiate interfaces in Java.

>>>>

This section illustrates how our programming style, supported by the
\mixinAnn{} annotation, enable powerful programming idioms based on multiple
inheritance and type refinements.  We propose a standard example:
\Q@Animal@s with a two-dimensional \Q@Point2D@ representing their
\Q@location@.  Some kinds of animals are \Q@Horses@ and \Q@Bird@s.
Birds can \Q@fly@, thus their locations need to be three-dimensional
\Q@Point3D@s (field type refinement).  Finally, we model \Q@Pegasus@
(one of the best-known creatures in Greek mythology) as a kind of
\Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s (multiple
inheritance). A simple class diagram illustrating the basic system is
given on the left side of Figure~\ref{fig:pegasus}.\footnote{Some
  research argues in favor of using subtyping for modeling taxonomies,
  other research argues against this practice, we do not wish to take
  sides in this argument, but to provide an engaging example.}

%\bruno{Should we provide a one sentence summary in the abstract of how much code
%is needed in Java (without Obj) vs the approach with CJ, for the Pegasus example?}
%\marco{Not in the abstract. But I think is good to have that number and to put in in the example section.
%I will set up the sentence and someone can compute the number.}

\subsection{Simple Multiple Inheritance with Default
  Methods}\label{sec:simple}

@@

\lstinputlisting[linerange=8-10]{../UseMixinLombok/src/pegasus/simple/lombok/Main.java}% APPLY:linerange=PEGASUS_LOMBOK
\bruno{I dislike the code formating here.}
\noindent The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a \texttt{Pegasus} object is instantiated as follows:

@@

The animal system modeled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}.
The example is still not sufficient to appreciate the advantages of IB
programming.
Lets now model the complete animal system where \Q@Animal@s includes a \Q@location@
representing their positions in space. We use 2D points to keep track of locations using coordinates.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Here we will illustrate how points are modelled with interfaces. In IB
we do not talk about state directly. Instead state is accessed and
manipulated using abstract methods.  The usual approach to model
points in Java is to use a class with fields for the coordinates. 
In Classless Java interfaces are used instead:

@@

\noindent The encoding over Java is now inconvenient: creating a new point object is cumbersome, even
with anonymous classes:

\begin{lstlisting}
Point2D p = new Point2D() {
  int x(){return 4;} public int y(){return 2;}  }
\end{lstlisting}

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with object interfaces}
This obvious ``constructor'' code can be automatically generated by the \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor. \mixin first looks at the abstract methods and detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{with-} methods in object interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable
data-structures. As such, it is
supported in by \mixin as \Q@with-@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D {
    int x();    int y(); // getters
     //with methods
    Point2D withX(int val);    Point2D withY(int val);
}
\end{lstlisting}

@@

\noindent If there is a large number of fields, \texttt{with} methods
will save programmers from writing large amounts of tedious code that
simply copies field values.
\begin{comment}
is expanded by \mixin into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provides a more uniform translation that can work also for mutable data structures, where setters are required.
}

@@

In summary, dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, and state is not directly
available to programmers. Instead programmers use methods, called
\emph{abstract state operations}, to interact with state.

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the \Q@Animal@ interface, the notion of points needs to
be \emph{refined}. Such kind of refinement is challenging
in typical class-based approaches. Fortunately, with object interfaces,
we are able to provide a simple and effective solution.

@@

\lstinputlisting[linerange=67-70]{../UseMixinLombok/src/pegasus/TestAnimal.java}% APPLY:linerange=BIRD
\bruno{Don't like the code format here either.}
%Interface \texttt{Point3D} extends \texttt{Point2D} with a new abstract method
%\texttt{int z()} (treated as a getter for member field \texttt{z}). Note that
%the return type of various methods (e.g. with- methods, getters) get refined
%either by covariant method overriding or automatically by our annotation
%processor. Besides \emph{with}, other methods (including \emph{clone},
%\emph{of}) also do type-refinements automatically.

@@

\paragraph{\Q@Point3D@ and properties updaters}
The \Q@Point3D@ interface is defined as follows:

@@

Property updaters never break class invariants, since they
internally call operations that were already deemed
safe by the programmer. For example a list object
would not offer a setter for its \texttt{size} field (which should be kept hidden), thus
a property updater would not attempt to set it.

@@

 \marcoT{Symmetrically,
 we could offer an imperative property updater
 that calls the setters instead of the withers.
 \Q@Point3D set(Point2D val)@.
 }\bruno{we could or we do? if we could then the table should not
   contain this operation; if we do the table is ok}

@@

\paragraph{Summary of operations in Classless Java} 
%Method bodies can only refer to other methods;
%Objects are (conceptually) closures and the state is 
%composed by captured local variables.}
%Methods interacting with state are called \emph{abstract state operations}.
Object interfaces provide support for
different types of abstract state operations: four field-based state
operations; and  two types of
general updating operations. Furthermore object interfaces support
direct object intantiation via \texttt{of} factory methods. 
Figure~\ref{fig:abstractstate} summarizes the six operations supported 
by \mixin. The field-based abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in more detail in Section~\ref{sec:dsls}.

\begin{figure*}
\centering
\saveSpaceFig
\begin{tabular}{|l|l|l|l|}
\hline
& \textbf{Operation}  & \textbf{Example}                  & \textbf{Description } \\ \hline
\multirow{4}{*}{\parbox{2.5cm}{State operations (for a field \texttt{x})}} & \textbf{``fields''/getters}        &   \Q@int x()@                  & Retrieves value from field \texttt{x}.          \\ \cline{2-4}
& {\bf withers}        &   \Q@Point2D withX(int val)@                & Clones
object; updates field \texttt{x} to \texttt{val}.             \\ \cline{2-4}
& \textbf{setters}        & \Q@void x(int val)@ & Sets the field
\texttt{x} to a  new value \texttt{val}.        \\ \cline{2-4}
& \textbf{fluent setters}        & \Q@Point2D x(int val)@ &Sets the field
\texttt{x} to \texttt{val} and returns \texttt{\this}.           \\ \hline
\multirow{3}{*}{Other operations} &
\textbf{factory methods} & 
%\begin{tabular}{l}
\Q@static Point2D of(int _x,int _y)@
%\end{tabular}
 & Factory method (generated).        \\
\cline{2-4}
& \textbf{functional updaters}        & \Q@Point3D with(Point2D val)@
& Update all matching fields in \texttt{val}.        \\
\cline{2-4}
& \textbf{imperative updaters}        & \Q@Point3D set(Point2D val)@ &
Set all matching fields in \texttt{val}.        \\ \hline
\end{tabular}

\caption{Abstract state operations for a field \texttt{x}, together with other operations, supported by the \mixin
  annotation. }

@@

\subsection{The Classless Java Approach}
Instead of writing a whole another class to provide the implementation for
\texttt{Point}, we annotate on interface \texttt{Point} directly with \mixin:

@@

With \CJ, we provide a Java annotation \mixin to provide default
implementations for various methods and a mechanism to instantiate
objects. \mixin annotation helps programmers to write less cumbersome code and
instantiate interfaces in Java.
